Supplement A: Different Response Distributions

### 2018-11-02 pkg/disttree/inst/draft/supplement/supplement_distributions.Rnw
final Z revision of main manuscript, adapting much of the discussion further touch-ups in remaining materials
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2375 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: a394dd57fb692b04399df40ba94b573455f36c1f

An alternative distribution with several similar properties but with
heavier tails is the logistic distribution. The PDF and CDF are given by
%
\begin{align}
f_{\mathrm{logistic}}(Y; \mu, \sigma) &= 
\frac{\exp{\left(-\frac{Y - \mu}{\sigma}\right)}}
{\sigma \cdot \left(1 + \exp{\left(-\frac{Y - \mu}{\sigma}\right)}\right)^2} \\
F_{\mathrm{logistic}}(Y; \mu, \sigma) &=
\frac{1}{1+ \exp{\left(-\frac{Y - \mu}{\sigma}\right)}}.
\end{align}
%
Both distributions have to be adapted in our case study to accommodate that
precipitation is non-negative with a point mass at zero (= no rain).
Two common strategies for this are to employ either \emph{left-censoring at zero}
or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
a truncated (positive) part \citep[see][for an overview of both techniques]{Winkelmann:2006}.
\begin{itemize}
\item \emph{Censored distribution:}
For a chosen distribution with PDF $f$ and CDF $F$ the log-likelihood function of the corresponding 
zero-censored distribution is
%
\begin{equation}
\ell_{\mathrm{cens}}(\mu, \sigma; Y) = 
\begin{cases}
\log\left\{f(Y; \mu, \sigma) \right\}, & \text{if } Y > 0 \\[0.2cm]
\log\left\{F(0; \mu, \sigma) \right\}, & \text{if } Y = 0
\end{cases}
\end{equation}
%
as stated in Equation~2.1.1 of the main manuscript for the Gaussian special case.
For positive values of $Y$ the censored version of the PDF
equals the original non-censored version. For $Y=0$ the censored density
function takes the values of the probability mass lying left of the 
censoring point. Therefore, the parameters $\mu$ and $\sigma$ control
both aspects, the point mass and the positive part of the distribution.

@@

Altogether it can be stated that for this application the left-censored at zero Gaussian
distribution seems to be an appropriate choice while the other two distributions are also
reasonable alternatives. However, they do not lead to an overall improvement in 
performance for this application.

@@

As a final consideration, we briefly remark that the distributional forest also allows
for a different hurdle model specification. Previously, we argued that for GAMLSS/EMOS
the log-likelihood of the hurdle model can be separated additively and hence the maximum
likelihood fit can be obtained by fitting two separate models for $\nu$ on one hand and
$\mu$ and $\sigma$ on the other. And therefore the same two-part approach was also adopted
for the distributional forests. However, an additional alternative is to fit a single
distributional forest based on the three-parameter log-likelihood from Equation~\ref{eq:hurdle}.
When all three parameters change simultaneously across the covariates, then this specification
can be expected to perform somewhat better. Whereas when different variables/effects drive
the parameters of the model, then the two-part approach implemented in this supplement
should perform somewhat better. In this particular case study, though, neither approach
performs clearly better as Figure~\ref{fig:box_crpss_forests} reveals. The figure depicts
CRPS skill scores for the 15~stations of the one-part hurdle models with the two-part
hurdle model as the reference. Clearly, both hurdle specifications perform very
similarly with skill scores being mostly below 1\%.

>>>>

An alternative distribution with several similar properties but with
heavier tails is the logistic distribution. The PDF and CDF are given by
%
\begin{align}
f_{\mathrm{logistic}}(Y; \mu, \sigma) &= 
\frac{\exp{\left(-\frac{Y - \mu}{\sigma}\right)}}
{\sigma \cdot \left(1 + \exp{\left(-\frac{Y - \mu}{\sigma}\right)}\right)^2} \\
F_{\mathrm{logistic}}(Y; \mu, \sigma) &=
\frac{1}{1+ \exp{\left(-\frac{Y - \mu}{\sigma}\right)}}.
\end{align}
%
Both distributions have to be adapted in our case study to accommodate that
precipitation is non-negative with a point mass at zero (= no rain).
Two common strategies for this are to employ either \emph{left-censoring at zero}
or a \emph{hurdle} (or two-part) approach, combining a binary (zero vs.\ greater) and
a truncated (positive) part (see \citealp{Long:1997} or \citealp{Winkelmann:2006}
for an overview of both techniques).
%
\begin{itemize}
\item \emph{Censored distribution:}
For a chosen distribution with PDF $f$ and CDF $F$ the log-likelihood function of the corresponding 
zero-censored distribution is
%
\begin{equation}
\ell_{\mathrm{cens}}(\mu, \sigma; Y) = 
\begin{cases}
\log\left\{f(Y; \mu, \sigma) \right\}, & \text{if } Y > 0 \\[0.2cm]
\log\left\{F(0; \mu, \sigma) \right\}, & \text{if } Y = 0
\end{cases}
\end{equation}
%
as stated in Equation~2.1.1 of the main manuscript for the Gaussian special case.
For positive values of $Y$ the censored version of the PDF
equals the original non-censored version. For $Y=0$ the censored density
function takes the values of the probability mass lying left of the 
censoring point. Therefore, the parameters $\mu$ and $\sigma$ control
both aspects, the point mass and the positive part of the distribution.

@@

Altogether it can be stated that for this application the zero-censored Gaussian
distribution seems to be an appropriate choice while the other two distributions are also
reasonable alternatives. However, they do not lead to a pronounced overall improvement in 
performance for this application.

@@

As a final consideration, we briefly remark that the distributional forest also allows
for a different hurdle model specification. Previously, we argued that for GAMLSS/EMOS
the log-likelihood of the hurdle model can be separated additively and hence the maximum
likelihood fit can be obtained by fitting two separate models for $\nu$ on one hand and
$\mu$ and $\sigma$ on the other. And therefore the same two-part approach was also adopted
for the distributional forests. However, an additional alternative is to fit a single
distributional forest based on the three-parameter log-likelihood from Equation~\ref{eq:hurdle}.
When all three parameters change simultaneously across the covariates, then this specification
can be expected to perform somewhat better. Whereas when different variables/effects drive
the different model parameters, then the two-part approach implemented in this supplement
should perform somewhat better. In this particular case study, though, neither approach
performs clearly better as Figure~\ref{fig:box_crpss_forests} reveals. The figure depicts
CRPS skill scores for the 15~stations of the one-part hurdle models with the two-part
hurdle model as the reference. Clearly, both hurdle specifications perform very
similarly with skill scores being mostly below 1\%.
